---
title: '8\. Worksheet: Among Site (Beta) Diversity -- Part 1'
author: "Atalanta Ritter; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this worksheet, we move beyond the investigation of within-site $\alpha$-diversity.
We will explore $\beta$-diversity, which is defined as the diversity that occurs among sites.
This requires that we examine the compositional similarity of assemblages that vary in space or time.

After completing this exercise you will know how to:

1. formally quantify $\beta$-diversity
2. visualize $\beta$-diversity with heatmaps, cluster analysis, and ordination
3. test hypotheses about $\beta$-diversity using multivariate statistics

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the  worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, it is *imperative* that you **push** this file to your GitHub repo, at whatever stage you are. Ths will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your '6.BetaDiversity' folder.
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file 
(**6.BetaDiversity_1_Worksheet.Rmd**) with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**6.BetaDiversity_1_Worksheet.pdf**).

The completed exercise is due on **Wednesday, February 1^st^, 2023 before 12:00 PM (noon)**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your "*/6.BetaDiversity*" folder, and
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list = ls())
getwd()
setwd("~/GitHub/QB2023_Ritter/2.Worksheets/6.BetaDiversity")
require(vegan)
```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, do the following:  

1. load the `doubs` dataset from the `ade4` package, and
2. explore the structure of the dataset.

```{r}
# note, please do not print the dataset when submitting
require(ade4)
data(doubs)
str(doubs, max.level = 1)
```

***Question 1***:  Describe some of the attributes of the `doubs` dataset.  

a.  How many objects are in `doubs`? 
b.  How many fish species are there in the `doubs` dataset?
c.  How many sites are in the `doubs` dataset?

> ***Answer 1a***:  There are 4 objects: env, fish, xy, and species.
> ***Answer 1b***:  27 fish species
> ***Answer 1c***:  30 sites

### Visualizing the Doubs River Dataset

***Question 2***: Answer the following questions based on the spatial patterns of richness (i.e., $\alpha$-diversity) and Brown Trout (*Salmo trutta*) abundance in the Doubs River.  

a.  How does fish richness vary along the sampled reach of the Doubs River?
b.  How does Brown Trout (*Salmo trutta*) abundance vary along the sampled reach of the Doubs River?
c.  What do these patterns say about the limitations of using richness when examining patterns of biodiversity?

> ***Answer 2a***:  Richness tends to increase as you go further down the Doubs River; upstream sites have lower richness, downstream sites have higher richness.
> ***Answer 2b***:  Brown trout tend to be more abundant in upstream sites as opposed to downstream sites.
> ***Answer 2c***: These patterns show how richness doesn't capture all of the information about a given ecological system. For instance, it doesn't capture the differences in evenness along the stream sites. Without looking at the brown trout abundance, we would not know that the upstream sites are predominantly made up of brown trout.

## 3) QUANTIFYING BETA-DIVERSITY

In the R code chunk below, do the following:  

1. write a function (`beta.w()`) to calculate Whittaker's $\beta$-diversity (i.e., $\beta_{w}$) that accepts a site-by-species matrix with optional arguments to specify pairwise turnover between two sites, and
2. use this function to analyze various aspects of $\beta$-diversity in the Doubs River.  

```{r}
beta.w <- function(site.by.species = "", sitenum1 = "", sitenum2 = "", 
                   pairwise = FALSE){
  # only if we specify pairwise as TRUE do this: 
  if (pairwise == TRUE){
    # as a check let's print an error if we do not provide needed arguments 
    if (sitenum1 == "" | sitenum2 == "") {
      print("Error: please specify sites to compare")
      return(NA)}
    
    site1 = site.by.species[sitenum1,]
    site2 = site.by.species[sitenum2,]
    site1 = subset(site1, select = site1 > 0)
    site2 = subset(site2, select = site2 > 0)
    gamma = union(colnames(site1), colnames(site2))
    s = length(gamma)
    a.bar = mean(c(specnumber(site1), specnumber(site2)))
    b.w = round(s/a.bar - 1,3)
    return(b.w)
  }
  else{
    SbyS.pa <- decostand(site.by.species, method = "pa")
    S <- ncol(SbyS.pa[,which(colSums(SbyS.pa) > 0)])
    a.bar <- mean(specnumber(SbyS.pa))
    b.w <- round(S/a.bar,3)
    return(b.w)
  }
}
# whittaker's beta diversity across all sites
beta.w(doubs$fish, pairwise = FALSE)
# beta diversity of site 1 vs. site 2
beta.w(doubs$fish, 1, 2, pairwise = TRUE)
# beta diversity of site 1 vs. site 10
beta.w(doubs$fish, 1, 10, pairwise = TRUE)
```

***Question 3***: Using your `beta.w()` function above, answer the following questions:

a.  Describe how local richness ($\alpha$) and turnover ($\beta$) contribute to regional ($\gamma$) fish diversity in the Doubs.
b.  Is the fish assemblage at site 1 more similar to the one at site 2 or site 10?
c.  Using your understanding of the equation $\beta_{w} = \gamma/\alpha$, how would your interpretation of $\beta$ change if we instead defined beta additively (i.e., $\beta = \gamma - \alpha$)?

> ***Answer 3a***: Beta is 2.16, which means that regional diversity is 2.16x higher than the average richness at each site. 
> ***Answer 3b***:  Site 1 is more similar to site 2. 
> ***Answer 3c***:  If $\beta$ were additive, it would become a measure of how many more species exist in the regional pool (gamma) than local sites (alpha). 

### The Resemblance Matrix
In order to quantify $\beta$-diversity for more than two samples, we need to introduce a new primary ecological data structure: the **Resemblance Matrix**. 

***Question 4***: How do incidence- and abundance-based metrics differ in their treatment of rare species?

> ***Answer 4***: Incidence-based metrics give rare species equal weight to common species, whereas abundance-based metrics take into account whether a species appeared many times vs. only a few times. 

In the R code chunk below, do the following:  

1. make a new object, `fish`, containing the fish abundance data for the Doubs River, 
2. remove any sites where no fish were observed (i.e., rows with sum of zero),
3. construct a resemblance matrix based on Sørensen's Similarity ("fish.ds"), and
4. construct a resemblance matrix based on Bray-Curtis Distance ("fish.db").  

```{r}
fish <- doubs$fish
fish <- fish[-8, ] # remove site 8 from data bc no fish were observed

# calculate sørenson
fish.ds <- vegdist(fish, method = "bray", binary = TRUE)
# calculate bray-curtis
fish.db <- vegdist(fish, method = "bray")

fish.ds
fish.db
```

***Question 5***: Using the distance matrices from above, answer the following questions:

a. Does the resemblance matrix (`fish.db`) represent similarity or dissimilarity? What information in the resemblance matrix led you to arrive at your answer? 
b. Compare the resemblance matrices (`fish.db` or `fish.ds`) you just created. How does the choice of the Sørensen or Bray-Curtis distance influence your interpretation of site (dis)similarity?

> ***Answer 5a***: It shows dissimilarity. Sites with numbers closer to 0 are very similar to each other whereas sites with numbers closer to 1 are dissimilar. So you're basically seeing a spectrum of similarity/dissimilarity across compared sites. 
> ***Answer 5b***: While the numbers are slightly different, my overall conclusion would be the same for both of them, that sites that are further down/upstream from each other tend to be more dissimilar than sites that are close together. 

## 4) VISUALIZING BETA-DIVERSITY

### A. Heatmaps

In the R code chunk below, do the following:  

1. define a color palette,
2. define the order of sites in the Doubs River, and
3. use the `levelplot()` function to create a heatmap of fish abundances in the Doubs River.

```{r}
require(viridis)
# define order of sites
order <- rev(attr(fish.db, "Labels"))
# plot heatmap
levelplot(as.matrix(fish.db)[, order], aspect = "iso", col.regions = turbo, xlab = "Doubs Site", 
          ylab = "Doubs Site", scales = list(cex =0.5, main = "Bray-Curtis Distance"))
```

### B.  Cluster Analysis

In the R code chunk below, do the following:  

1. perform a cluster analysis using Ward's Clustering, and
2. plot your cluster analysis (use either `hclust` or `heatmap.2`). 

```{r} 
# ward's clustering
doubs.ward <- hclust(fish.db, method = "ward.D2")
# plot cluster analysis
par(mar = c(1,5,2,2) + 0.1)
gplots::heatmap.2(as.matrix(fish), 
                  distfun = function(x) vegdist(x, method = "bray"),
                  hclustfun = function(x) hclust(x, method = "ward.D2"),
                  col = viridis, trace = "none", density.info = "none")
```

***Question 6***: Based on cluster analyses and the introductory plots that we generated after loading the data, develop an ecological hypothesis for fish diversity the `doubs` data set? 

> ***Answer 6***: I hypothesize that fish diversity increases as you go further downstream because there are more resources available and more habitat variation to accomodate different species of fish. 

### C.  Ordination

### Principal Coordinates Analysis (PCoA)

In the R code chunk below, do the following:  

1. perform a Principal Coordinates Analysis to visualize beta-diversity
2. calculate the variation explained by the first three axes in your ordination
3. plot the PCoA ordination,
4. label the sites as points using the Doubs River site number, and
5. identify influential species and add species coordinates to PCoA plot.

```{r}
# 1. perform a Principal Coordinates Analysis to visualize beta-diversity
fish.pcoa <- cmdscale(fish.db, eig = TRUE, k = 3)
# 2. calculate the variation explained by the first three axes in your ordination
explainvar1 <- round(fish.pcoa$eig[1] / sum(fish.pcoa$eig), 3) * 100
explainvar2 <- round(fish.pcoa$eig[2] / sum(fish.pcoa$eig), 3) * 100
explainvar3 <- round(fish.pcoa$eig[3] / sum(fish.pcoa$eig), 3) * 100
# 3. make PCOA plot
# define plot parameters
par(mar = c(5, 5, 1, 2) + 0.1)
# initiate plot
plot(fish.pcoa$points[ ,1], fish.pcoa$points[ , 2], ylim = c(-0.2, 0.7), 
     xlab = paste("PCoA 1 (", explainvar1, "%)", sep = ""),
     ylab = paste("PCoA 2 (", explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, 
     cex.axis = 1.2, axes = FALSE) 
# add axes
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)
# add points and labels
points(fish.pcoa$points[ , 1], fish.pcoa$points[ ,2],
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(fish.pcoa$points[ ,1], fish.pcoa$points[ ,2],
     labels = row.names(fish.pcoa$points))
# 5. identify influential species and add species coordinates to PCoA plot
fishREL <- fish
  for(i in 1:nrow(fish)){
    fishREL[i, ] = fish[i, ] / sum(fish[i, ])
  }
library(BiodiversityR)
fish.pcoa <- add.spec.scores(fish.pcoa, fishREL, method = "pcoa.scores")
text(fish.pcoa$cproj[ ,1], fish.pcoa$cproj[ ,2], 
     labels = row.names(fish.pcoa$cproj), col = "black")
```

In the R code chunk below, do the following:  

1. identify influential species based on correlations along each PCoA axis (use a cutoff of 0.70), and
2. use a permutation test (999 permutations) to test the correlations of each species along each axis.

```{r}
spe.corr <- add.spec.scores(fish.pcoa, fishREL, method = "cor.scores")$cproj
corrcut <- 0.7  # user defined cutoff
imp.spp <- spe.corr[abs(spe.corr[, 1]) <= corrcut | abs(spe.corr[, 2]) <= corrcut, ]
#permutation test for species abundances across axes
fit <- envfit(fish.pcoa, fishREL, perm = 999)
```

***Question 7***: Address the following questions about the ordination results of the `doubs` data set:

a.  Describe the grouping of sites in the Doubs River based on fish community composition.
b.  Generate a hypothesis about which fish species are potential indicators of river quality.

> ***Answer 7a***: (Brown trout groups with the upstream sites, which aligns with the abundance data for that species.) 
> ***Answer 7b***: I think brown trout, minnow, and stone loach could be indicators of river quality because they cluster with the upstream sites, which are assumedly less polluted. 


## SYNTHESIS

Load the dataset from that you and your partner are using for the team project. 
Use one of the tools introduced in the beta diversity module to visualize your data. 
Describe any interesting patterns and identify a hypothesis is relevant to the principles of biodiversity. 

```{r}
# loading in dataset
seabirds <- read.csv("/Users/Atalanta/Downloads/Data_DRYAD_Seabirds.csv")
seabirds.lite <- seabirds[-c(37,59, 75,76,78,79,82,83,112,121,135,165,172,177,190,198,259,280,307,311,312) ,2:5] 
# remove xtra columns and any rows with all 0s 
# cluster analysis
par(mar = c(1,5,2,2) + 0.1)
gplots::heatmap.2(as.matrix(seabirds.lite), 
                  distfun = function(x) vegdist(x, method = "bray"),
                  hclustfun = function(x) hclust(x, method = "ward.D2"),
                  col = viridis, trace = "none", density.info = "none")
```
> Overall, sites seem to be pretty similar to each other in terms of each species abundance. However, there appears to be a cluster of sites including site 70 that has an uncharacteristically high abundance of shearwaters. Based on this, I hypothesize that shearwaters congregate in very specific locales rather than disperse. 
